---
title: "play4YES: probability illustrations with playing cards"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{play4YES: probability illustrations with playing cards}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# A deck of cards as an R vector

```{r do1}
library(play4YES)
set.seed(1234)  # reproducible
deck = build_deck()
deck
```

# Is it a fair deck?

```{r do2}
table(suits(deck), faces(deck))
```

Make a biased deck:
```{r do3}
bdeck = deck
bdeck[3] = bdeck[18]
table(suits(bdeck), faces(bdeck))
```

# Fairness assessed by top card in a series of deals

We'll use the built-in function `sample` to
simulate shuffles of the deck.  `top_card` gives
the card at top of deck.

```{r redef}
shuffle = sample
top_card = function(x) x[1]
t1 = top_card(shuffle(deck))
t1
suits(t1)
```

## Using the fair deck

We'll take the top cards for each of 100 shuffles and examine the suit.
```{r do4}
table( replicate(100, suits(top_card(shuffle(deck)))) )
```

We should see about 25 for each suit.  Are we close enough?

## Using the biased deck

```{r do5}
table( replicate(100, suits(top_card(shuffle(bdeck)))) )
```

## Increasing the "sample size"

Variability can mask structural bias.  If we create statistics
based on more observations, we can start to unearth it.

```{r do6}
table( replicate(10000, suits(top_card(shuffle(deck)))) )
table( replicate(10000, suits(top_card(shuffle(bdeck)))) )
```

# A finer-grained problem: the face, not the suit

There are 4 different suits, but 13 different faces.  Is
the bias discernible with this procedure?

```{r do7a}
table( replicate(100, faces(top_card(shuffle(deck)))) )
table( replicate(100, faces(top_card(shuffle(bdeck)))) )
```

If we make 10000 shuffles, we expect each face
to appear 10000/13 = 769.230.... times.

```{r do7}
tfair <- table( replicate(10000, faces(top_card(shuffle(deck)))) )
tfair
tbias <- table( replicate(10000, faces(top_card(shuffle(bdeck)))) )
tbias
```

We can plot the observed frequencies of appearance of each face.

```{r do9}
library(beeswarm)
beeswarm(list(fair=tfair,biased=tbias), ylab="Top card face frequencies")
```

With the biased deck, one face has depleted frequency, one enhanced.
